---
title: "p8105_hw2_jc6961.Rmd"
author: "Junze Cao"
date: "2026-10-05"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
```

## Problem 1

### Import and clean the subway data

```{r subway_import, message = FALSE}
```{r}
subway_df =
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route1 = col_character(),
      Route2 = col_character(),
      Route3 = col_character(),
      Route4 = col_character(),
      Route5 = col_character(),
      Route6 = col_character(),
      Route7 = col_character(),
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE))

head(subway_df)
```

The cleaned subway dataset has `r nrow(subway_df)` rows and
`r ncol(subway_df)` columns. Each row describes a subway entrance or exit.
The retained variables describe the line, station name, station latitude
and longitude, routes served, entry permission, vending availability,
entrance type, and ADA compliance. I cleaned the column names, retained
only the requested variables, and converted entry permission from YES/NO
to TRUE/FALSE. I imported all route columns as character variables because
train names may contain letters or numbers.

### Count distinct stations and examine entry permission

```{r subway_summary}
```{r}
station_df =
  subway_df |>
  distinct(line, station_name)

ada_station_df =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name)

no_vending_df =
  subway_df |>
  filter(vending == "NO")
```

There are `r nrow(station_df)` distinct stations, identified by the
combination of line and station name. Of these,
`r nrow(ada_station_df)` are ADA compliant. Counting distinct stations
avoids counting a station repeatedly because it has multiple entrances.

Among the `r nrow(no_vending_df)` entrances/exits without vending,
`r nrow(filter(no_vending_df, entry == TRUE))` allow entry.
The proportion allowing entry is
`r round(mean(pull(no_vending_df, entry)), 4)`, or
`r round(mean(pull(no_vending_df, entry)) * 100, 2)`%.
For a logical variable, the mean equals the proportion of TRUE values.

### Reshape the route columns

```{r subway_routes}
```{r}
subway_tidy_df =
  subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |>
  drop_na(route_name) |>
  mutate(route_number = as.integer(route_number))

a_station_df =
  subway_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name)

a_ada_station_df =
  subway_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name)

head(subway_tidy_df)
```

The reshaped dataset has separate variables for the original route slot
(`route_number`) and the train designation (`route_name`). Each row
represents an entrance/exit and one of the routes it serves; unused route
slots are removed. There are `r nrow(a_station_df)` distinct stations
serving the A train, of which `r nrow(a_ada_station_df)` are ADA compliant.

## Problem 2

### Import and combine the Trash Wheel data

```{r trash_import, message = FALSE}
mr_trash_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N710") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel")

professor_trash_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M135") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel")

gwynnda_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L352") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda")

trash_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df) |>
  relocate(trash_wheel, dumpster)

head(trash_df)
```

I used the sheet and range arguments to select the appropriate data
rectangles and exclude notes and unrelated columns. The ranges begin at
the column-header row and include a final totals row, which is removed
using `drop_na(dumpster)`. I cleaned variable names, converted year to an
integer consistently across sheets, and rounded Mr. Trash Wheel's
sports-ball counts before converting them to integers. I added a variable
identifying each wheel before combining the three datasets with
`bind_rows()`.

The combined tidy dataset has `r nrow(trash_df)` observations and
`r ncol(trash_df)` variables, with `r nrow(mr_trash_df)` observations from
Mr. Trash Wheel, `r nrow(professor_trash_df)` from Professor Trash Wheel,
and `r nrow(gwynnda_df)` from Gwynnda. Each observation represents a
single dumpster collection from a particular wheel. Important variables
include collection date, weight in tons, volume in cubic yards, and counts
of plastic bottles, cigarette butts, and other trash. Dumpster numbers are
only unique within a wheel. Sports-ball counts are only available for
Mr. Trash Wheel, and Gwynnda has no glass-bottle column. Combining the
sheets leaves these unavailable values as NA rather than treating them
as zero.

### Calculate the requested totals

```{r trash_totals}
professor_weight_df =
  trash_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  select(weight_tons) |>
  drop_na(weight_tons)

gwynnda_june_df =
  trash_df |>
  filter(trash_wheel == "Gwynnda", year == 2022, month == "June") |>
  select(cigarette_butts) |>
  drop_na(cigarette_butts)

# Total = mean * number of non-missing observations.
professor_total_weight =
  mean(pull(professor_weight_df, weight_tons)) * nrow(professor_weight_df)

gwynnda_june_cigarettes =
  mean(pull(gwynnda_june_df, cigarette_butts)) * nrow(gwynnda_june_df)
```

For the available data, Professor Trash Wheel collected a total of
`r round(professor_total_weight, 2)` tons of trash, based on
`r nrow(professor_weight_df)` non-missing weight records.
Gwynnda collected `r round(gwynnda_june_cigarettes)` cigarette butts
in June 2022, based on `r nrow(gwynnda_june_df)` non-missing records.
The totals are calculated as the mean multiplied by the number of
non-missing observations; the observation counts are computed from the
data rather than entered manually.

