---
title: "p8105_hw2_jc6961.Rmd"
author: "Junze Cao"
date: "2026-10-05"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
```

## Problem 1

### Import and clean the subway data

```{r warning=FALSE}
subway_df =
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route1 = col_character(),
      Route2 = col_character(),
      Route3 = col_character(),
      Route4 = col_character(),
      Route5 = col_character(),
      Route6 = col_character(),
      Route7 = col_character(),
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE))

head(subway_df)
```

The cleaned subway dataset has `r nrow(subway_df)` rows and `r ncol(subway_df)` columns. It contains line and station information, station location, routes served, entry, vending, entrance type, and ADA status. I cleaned the variable names, kept the requested variables, and changed `entry` from YES/NO to TRUE/FALSE. The data are not tidy because routes are stored in multiple columns.

### Count distinct stations and examine entry permission

```{r}
station_df =
  subway_df |>
  distinct(line, station_name)

ada_station_df =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name)

no_vending_df =
  subway_df |>
  filter(vending == "NO")
```

There are `r nrow(station_df)` distinct stations, and `r nrow(ada_station_df)` are ADA compliant.

Among entrances/exits without vending, `r round(mean(pull(no_vending_df, entry)) * 100, 2)`% allow entry.


### Reshape the route columns

```{r}
subway_tidy_df =
  subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |>
  drop_na(route_name) |>
  mutate(route_number = as.integer(route_number))

a_station_df =
  subway_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name)

a_ada_station_df =
  subway_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name)

head(subway_tidy_df)
```

After reshaping, route number and route name are separate variables. There are `r nrow(a_station_df)` distinct stations serving the A train, and `r nrow(a_ada_station_df)` are ADA compliant.


## Problem 2

### Import and combine the Trash Wheel data

```{r}
mr_trash_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N710") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel")

professor_trash_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M135") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel")

gwynnda_df =
  readxl::read_excel(
    "./data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L352") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda")

trash_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df) |>
  relocate(trash_wheel, dumpster)

head(trash_df)
```

I imported the three Trash Wheel sheets using the appropriate sheet and range arguments to exclude notes and non-data entries. I cleaned the variable names, removed rows without dumpster data, and added a variable identifying each Trash Wheel. For Mr. Trash Wheel, sports balls were rounded and converted to integers.

The combined tidy dataset has `r nrow(trash_df)` observations and `r ncol(trash_df)` variables. Each row represents one dumpster collection. Key variables include collection date, weight, volume, plastic bottles, and cigarette butts.


### Calculate the requested totals

```{r}
professor_weight_df =
  trash_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  select(weight_tons) |>
  drop_na(weight_tons)

gwynnda_june_df =
  trash_df |>
  filter(trash_wheel == "Gwynnda", year == 2022, month == "June") |>
  select(cigarette_butts) |>
  drop_na(cigarette_butts)

professor_total_weight =
  sum(pull(professor_weight_df, weight_tons))

gwynnda_june_cigarettes =
  sum(pull(gwynnda_june_df, cigarette_butts))
```

Professor Trash Wheel collected `r round(professor_total_weight, 2)` tons of trash in the available data.

Gwynnda collected `r format(round(gwynnda_june_cigarettes), scientific = FALSE, big.mark = ",")` cigarette butts in June 2022.


## Problem 3

### Import and clean the baseline demographics

```{r}
baseline_raw_df =
  read_csv(
    "./data/MCI_baseline.csv",
    skip = 1,
    na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"),
    sex = as.factor(sex),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"),
    apoe4 = as.factor(apoe4))

baseline_excluded_df =
  baseline_raw_df |>
  filter(age_at_onset <= current_age)

baseline_df =
  baseline_raw_df |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

head(baseline_df)
```

I skipped the first descriptive row, cleaned the variable names, treated missing-value codes as `NA`, and converted sex and APOE4 status from numeric codes to factors. Participants who already had MCI at baseline were removed. The cleaned data are tidy, with one row per participant.

After excluding `r nrow(baseline_excluded_df)` participants who had MCI at baseline, the study included `r nrow(baseline_df)` participants.


```{r}
mci_developed_df =
  baseline_df |>
  filter(!is.na(age_at_onset))

women_df =
  baseline_df |>
  filter(sex == "female")
```

Among the eligible participants, `r nrow(mci_developed_df)` developed MCI during follow-up. The average baseline age is `r round(mean(pull(baseline_df, current_age), na.rm = TRUE), 2)` years. Among women, `r round(mean(pull(women_df, apoe4) == "carrier") * 100, 1)`% are APOE4 carriers.


### Import and tidy the longitudinal biomarker data

```{r}
amyloid_df =
  read_csv(
    "./data/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na", "", ".")) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid_ratio") |>
  mutate(
    time = case_match(
      time,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8))

head(amyloid_df)
```

For the biomarker data, I skipped the descriptive first row, cleaned the variable names, renamed the participant ID to match the baseline data, and treated the missing-value codes as `NA`. I reshaped the repeated amyloid measurements into long format, with time recorded as years since baseline.

The tidy biomarker dataset contains `r nrow(distinct(amyloid_df, id))` participants and `r nrow(distinct(amyloid_df, time))` scheduled measurement times. Each row represents one participant at one time point. There are `r nrow(filter(amyloid_df, is.na(amyloid_ratio)))` missing amyloid-ratio values.

### Check for unmatched participants

```{r}
baseline_only_df =
  baseline_df |>
  filter(!(id %in% pull(amyloid_df, id))) |>
  select(id)

amyloid_only_df =
  amyloid_df |>
  filter(!(id %in% pull(baseline_df, id))) |>
  distinct(id)

amyloid_only_raw_df =
  amyloid_df |>
  filter(!(id %in% pull(baseline_raw_df, id))) |>
  distinct(id)
```

There are `r nrow(baseline_only_df)` participants in the eligible baseline dataset who do not appear in the biomarker dataset, and `r nrow(amyloid_only_df)` participants in the biomarker dataset who do not appear in the eligible baseline dataset.

Of the biomarker-only participants, `r nrow(amyloid_only_raw_df)` (IDs `r pull(amyloid_only_raw_df, id)`) are not in the original baseline file. The remaining `r nrow(amyloid_only_df) - nrow(amyloid_only_raw_df)` were removed from the baseline data because they already had MCI at baseline.


### Combine and export the data

```{r}
baseline_matched_df =
  baseline_df |>
  filter(id %in% pull(amyloid_df, id))

mci_df =
  left_join(baseline_matched_df, amyloid_df, by = "id") |>
  arrange(id, time)

write_csv(mci_df, "./data/mci_combined.csv")

head(mci_df)
```

After keeping only participants found in both datasets, the final dataset contains `r nrow(distinct(mci_df, id))` participants, `r nrow(mci_df)` rows, and `r ncol(mci_df)` columns. Each row represents one participant at one scheduled measurement time, with baseline demographic information repeated across visits. The final dataset was exported to `./data/mci_combined.csv`.
