---
title: "p8105_hw2_jc6961.Rmd"
author: "Junze Cao"
date: "2026-10-05"
output: github_document
---

```{r message=FALSE}
library(tidyverse)
```

## Problem 1

### Import and clean the subway data

```{r subway_import, message = FALSE}
subway_df =
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route1 = col_character(),
      Route2 = col_character(),
      Route3 = col_character(),
      Route4 = col_character(),
      Route5 = col_character(),
      Route6 = col_character(),
      Route7 = col_character(),
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE))

head(subway_df)
```

The cleaned subway dataset has `r nrow(subway_df)` rows and
`r ncol(subway_df)` columns. Each row describes a subway entrance or exit.
The retained variables describe the line, station name, station latitude
and longitude, routes served, entry permission, vending availability,
entrance type, and ADA compliance. I cleaned the column names, retained
only the requested variables, and converted entry permission from YES/NO
to TRUE/FALSE. I imported all route columns as character variables because
train names may contain letters or numbers.

### Count distinct stations and examine entry permission

```{r subway_summary}
station_df =
  subway_df |>
  distinct(line, station_name)

ada_station_df =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name)

no_vending_df =
  subway_df |>
  filter(vending == "NO")
```

There are `r nrow(station_df)` distinct stations, identified by the
combination of line and station name. Of these,
`r nrow(ada_station_df)` are ADA compliant. Counting distinct stations
avoids counting a station repeatedly because it has multiple entrances.

Among the `r nrow(no_vending_df)` entrances/exits without vending,
`r nrow(filter(no_vending_df, entry == TRUE))` allow entry.
The proportion allowing entry is
`r round(mean(pull(no_vending_df, entry)), 4)`, or
`r round(mean(pull(no_vending_df, entry)) * 100, 2)`%.
For a logical variable, the mean equals the proportion of TRUE values.

### Reshape the route columns

```{r subway_routes}
subway_tidy_df =
  subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |>
  drop_na(route_name) |>
  mutate(route_number = as.integer(route_number))

a_station_df =
  subway_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name)

a_ada_station_df =
  subway_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name)

head(subway_tidy_df)
```

The reshaped dataset has separate variables for the original route slot
(`route_number`) and the train designation (`route_name`). Each row
represents an entrance/exit and one of the routes it serves; unused route
slots are removed. There are `r nrow(a_station_df)` distinct stations
serving the A train, of which `r nrow(a_ada_station_df)` are ADA compliant.
